---
title: "PE-PSet1"
output: pdf_document
author: Drazzel Feliu - 12174100
---

```{r setup, results='hide', warning=FALSE, message=FALSE,include=FALSE}
# Clear Workspace
rm(list=ls())

# Load Libraries
library(haven)
library(reshape)
library(tidyverse)
library(magrittr)
library(survival)
```

For this assignment, provide a write-up where you answer the questions below, selectively cutting and pasting output where needed. Be concise in your write-up; excess wordiness will be penalized. Also submit a log file that includes commands and results for your entire analysis.

# Study motivation
Developing and middle-income countries increasingly provide health coverage to their residents. Mexico has become a leading example. Prior to 2004, roughly half the Mexican population had no health insurance. Health coverage was available through the social security system for salaried employees and their families, but the roughly 50 million Mexicans outside the salaried sector had no insurance.

In 2004, Mexico introduced Seguro Popular. Seguro Popular is a voluntary, non- contributory health insurance program for residents without coverage from the social security system. One of its explicit goals is to provide covered households with financial protection by reducing their incidence of catastrophic out-of-pocket health spending.

# The Seguro Popular field experiment
The experiment involved a paired-cluster randomization design, by which 100 largely rural “health clusters” were formed into 50 pairs on the basis of similarities in demographic characteristics and healthcare infrastructure. Health clusters are defined as a healthcare facility and the population living within its catchment area. Within matched pairs of health clusters, one was randomly assigned to treatment, with the other serving as control. Treatment involved publicity campaigns designed to encourage local residents to enroll. Extensive information about the initial design of the experiment has been published by King et al (2009).

Pre-intervention baseline data were collected from participant households in 2005. Post-intervention follow-up data were collected in 2006, roughly 10 months after the intervention began. The 2005 survey included 32,515 households.

# Questions

## (1)
Did Seguro Popular reduce extreme expenditures on health? 

## (2)
Does it matter how we define “extreme expenditures?”

# Data
Data for this assignment are in the file SP_dataset.dta on Canvas.

```{r, echo=FALSE}
# Load dataset
data <- read_dta("SP_dataset.dta")
# identify data labels
datalabels <- sapply(data, function (x) attr(x, "label"))
```

# Analysis

## (1) 
Inspect the data. Are there truly 100 clusters in 50 pairs? Does each pair have a treatment and a control cluster? What is the modal level of schooling in the sample? The modal marital status? What share of sample households were participating in Oportunidades, Mexico’s conditional cash transfer program? What is mean non- health expenditure (data are in Mexican pesos, where 12 pesos = approx. 1 dollar)? Does this make sense?

```{r, results='hide'}
# data cleaning
# check data class & view summary
sapply(data, class)
summary(data)

# assuming that the data is missing completely at random, all NA values across all observations must be dropped
data <- na.omit(data)

# view data summary
summary(data)
```

## Answer
There are exactly 100 clusters and 50 cluster pairs. Each pair does indeed have a treatment and a control cluster. The modal level of education is elementary education (11484). The modal marital status is married (13398). The share of sample households participating in Opurtunidades is (11192). The mean non-health expenditure in 2005 is 9829.203 in pesos and 819.1002 in US dollars. The mean non-health expenditure in 2006 is 9737.579 in pesos and 811.4649 in US dollars.
```{r}
# number of clusters
length(unique(data$cluster))
# number of cluster pairs
length(unique(data$clust_pair))
# number of treated and untreated in each cluster pair
cast(data, clust_pair ~ treatment)

# modal level of schooling
colSums(data[17:22])
# modal marital status
colSums(data[27:32])

# of households participating in Oportunidades
sum(data$opor_05)

# mean non-health expenditure in 2005 in pesos
mean(data$allbut_05)
# mean non-health expenditure in 2005 in dollars
(mean(data$allbut_05))/12

# mean non-health expenditure in 2006 in pesos
mean(data$allbut_06)
# mean non-health expenditure in 2006 in dollars
(mean(data$allbut_06))/12
```

## (2) 
Which treatment parameters could be identified by these data?

## Answer
The treatment parameters that can be identified by the data are the varying types of expenditures that differ between 2005 and 2006 (food expenditure, all non-medical expenditure, medical expenditure). Using an instrumental variable, assuming that monotonicity holds, we can identify the impact of treatment on the compliers thereby producing the local ATE. In this case, identifying those who would respond to treatment is a meaningful group worthy of being analyzed as this program is meant to identify those at the margins who would benefit from being insured against catastrophic out-of-pocket health spending.

## (3) 
Assess the adequacy of random assignment. You will need to think about how to use Stata’s reg command to test for equal means among pre-intervention variables within cluster pairs. How important is it to account for dependence within clusters? What is the bottom line? Was random assignment properly executed?

## Answer

```{r}
balance <- lm(data = data,treatment ~ sex + age + marstat + hhsize + food_yr_05 + allbut_05 + oop_yr3_05+ nkid_05 + nadult_05 + headwomen_05 + edu_info_051 + edu_info_052+ edu_info_053+ edu_info_054+ edu_info_055+ edu_info_056 + insp2005 + opor_05, cluster(data$clust_pair))
summary(balance)

```

Random assignment seems to have been properly executed across the board with the exception of one variable (oop_yr3_05). It's important to account for dependence within clusters to ensure that the treatment and control groups are reflective of each other within the clusters so that we can draw conclusions broadly about whether or not treatment was effective.

## (4) 
Construct two variables relating health spending to the household’s total income. The first is the health budget share, equal to health spending divided by total spending. The second is the adjusted health budget share, equal to health spending divided by disposable income, where disposable income is defined as total spending less food expenditures.

## Answer

```{r}
# create health budget share in 2005
data$hbshare05 <- data$oop_yr3_05/(data$allbut_05+data$oop_yr3_05)
# create health budget share in 2006
data$hbshare06 <- data$oop_yr3_06/(data$allbut_06+data$oop_yr3_06)

# create adjusted health budget share in 2006
data$adjhbshare05 <- data$oop_yr3_05/(data$allbut_05+data$oop_yr3_05-data$food_yr_05)
# create adjusted health budget share in 2006
data$adjhbshare06 <- data$oop_yr3_06/(data$allbut_06+data$oop_yr3_06-data$food_yr_06)
```

## (5) 
On average, what share of total expenditures does health spending account for? What are the 75th and 90th percentiles of the budget share distribution? What are the corresponding figures as a share of disposable income rather than total expenditures?

## Answer

Mean health budget shares (2005 & 2006) and mean adjusted health budget shares (2005 & 2006).

```{r}
# average health budget share in 2005
mean(data$hbshare05, na.rm = TRUE)
# average health budget share in 2006
mean(data$hbshare06, na.rm = TRUE)
# average adjusted health budget share in 2005
mean(data$adjhbshare05, na.rm = TRUE)
# average adjusted health budget share in 2006
mean(data$adjhbshare06, na.rm = TRUE)
```

75th percentile for health budget share (2005 & 2006) and adjusted health budget shares (2005 & 2006)

```{r}
# 75th percentile for health budget share in 2005
quantile(data$hbshare05, .75, na.rm = TRUE)
# 75th percentile for health budget share in 2006
quantile(data$hbshare06, .75, na.rm = TRUE)
# 90th percentile for health budget share in 2005
quantile(data$hbshare05, .90, na.rm = TRUE)
# 90th percentile for health budget share in 2005
quantile(data$hbshare06, .90, na.rm = TRUE)
```

90th percentile for health budget share (2005 & 2006) and adjusted health budget shares (2005 & 2006)

```{r}
# 75th percentile for adjusted health budget share in 2005
quantile(data$adjhbshare05, .75, na.rm = TRUE)
# 75th percentile for adjusted health budget share in 2006
quantile(data$adjhbshare06, .75, na.rm = TRUE)
# 90th percentile for adjusted health budget share in 2005
quantile(data$adjhbshare05, .90, na.rm = TRUE)
# 90th percentile for adjusted health budget share in 2006
quantile(data$adjhbshare06, .90, na.rm = TRUE)
```

## (6) 
Construct two measures of extreme health spending: (i) a dummy equal to one if the household’s health budget share exceeds 20 percent; and (ii) a dummy equal to one if the household spends more than 30 percent of its disposable income on health. How are these measures related?

## Answer

```{r}
data$extreme20dummy05 <- ifelse(data$hbshare05 > .20,1,0)
data$extreme20dummy06 <- ifelse(data$hbshare06 > .20,1,0)
  
data$extreme30dummy05 <- ifelse(data$adjhbshare05 > .30,1,0)
data$extreme30dummy06 <- ifelse(data$adjhbshare06 > .30,1,0)
```

These measures are both reflections of how much of the household budget is consumed by medical spending. Given that the second dummy variable has both a higher threshhold and is made up of discretionary spending, it is likely more sensitive to the impact of health spending than that first dummy variable, as this one includes the fixed costs of food as part of its calculation.

## (7) 
Estimate the ITT effects of the intervention on extreme health spending. First, just regress the extreme health spending measures on the treatment dummy. Next, add cluster-pair dummies. In both cases, calculate standard errors that account for dependence among households within clusters. What happens to the coefficients when you add the cluster-pair dummies? To their standard errors? Why?

## Answer 

```{r}
# regress extreme health budget share spending in 2006 on treatment
itt <- lm(extreme20dummy06 ~ treatment, data = data)
sumitt <- summary(itt)
sumitt
# regress extreme health budget share spending on treatment with cluster dummies
itt2 <- lm(extreme20dummy06 ~ treatment + factor(clust_pair), data = data)
sumitt2 <- summary(itt2)
sumitt2
```

The coefficients in both sets of regressions lose precision. Regressing extreme health budget share on only the treatment dummy provides a coefficient of $-0.01495$ with a standard error of $0.00282$. When including the cluster pair dummy, the treatment coefficient becomes $-0.01479$. with a standard error of $0.00290$. This follows logically, as any instrumental variable, no matter how effective it is, will diminish the precision of OLS estimators. Our instrument variable's F-statistic is less than 10, indicating the instrument is weak. 

```{r}
# regress adjusted extreme health budget share spending in 2006 on treatment
itt3 <- lm(extreme30dummy06 ~ treatment, data = data)
sumitt3 <- summary(itt3)
sumitt3
# regress adjusted extreme health budget share spending on treatment with cluster dummies
itt4 <- lm(extreme30dummy06 ~ treatment + factor(clust_pair), data = data)
sumitt4 <- summary(itt4)
sumitt4
```

Our second set of regressions shows a similar relationship to our initial set of regressions. Coefficients lose precision. And our standard errors grow. Our F-statistic is less than 10, again indicating a weak instrument.

## (8) 
Now adjust for baseline characteristics. Again, do this with and without the cluster-pair dummies in the regression. What happens to the estimated treatment effects? To their standard errors?

## Answer

```{r}
itt5 <- lm(data = data, extreme20dummy06 ~ treatment + sex + age + hhsize + headwomen_05)
sumitt5 <- summary(itt5)
sumitt5

itt6 <- lm(data = data, extreme20dummy06 ~ treatment + sex + age + hhsize + headwomen_05 + factor(clust_pair))
sumitt6 <- summary(itt6)
sumitt6
```

We've chosen the following variables to be our baseline: sex, age, hhsize, & headwomen_05. As with previous examples, we note similar differences between the coefficients of our baseline values in the regression without dummy variables for cluster pair and with. The coefficients become less precise. The standard errors increase.

## (9) 
Are any of your conclusions sensitive to the way that you measure extreme health expenditures?

## Answer

I do not believe so. The coefficients for the health share are larger than the coefficients for the adjusted health share in either regression (dummy variables or not). But they do not exhibit what might be a statistically significant difference from each other. As such they both seem to be similarly effective parameters.

## (10)
Now calculate the Wald estimator. How do you interpret it? What additional assumptions do you need to invoke for these estimates to be valid?

## Answer

```{r}
firststage <- lm(data = data, insp2006 ~ treatment)
fs <- summary(firststage)
fs

insp2006hat <- firststage$fitted.values
secondstage <- lm(data = data, extreme20dummy06 ~ insp2006hat)
ss <- summary(secondstage)
ss

ss$coefficients[2]

```

In order to calculate the Wald estimator, we need several assumptions to hold. We need there to be no interference or that there exist no spillover from one sample to another. We require that Z has no independent effect on Y, except through D. We need there to be a significant difference in D, given either value of Z. And we also need the monotonicity assumption, highlight prior, to hold.
